---
title: "The Collection and Analysis of Zooplankton"
author: "Claire Estey"
date: "`r Sys.Date()`"
output: html_document
bibliography: "bib_file.bib"
---

### Notes (not to be included)

put this in a folder titled final or something and have it render as an html_ file_document, leave README as .Rmd

try to change date to system date in yaml header 
can also change html style (tufte?)
due?
integrate equations 
bibliography
needs to be reproducible - just share with Dr. Paez (not a collaborator)

practice ref: [@arnold2023quat]

### Install Packages

- tidyverse
- here

### *Introduction*

  Benzalkonium chloride (BAC) is a potent antimicrobial currently listed as the number one active ingredient in disinfectants approved by the Government of Canada for use against COVID-19 [@marteinson2022increased]. Consequently, BAC use has increased in industrial, clinical, and household settings since the onset of the pandemic [@marteinson2022increased; @mohapatra2023quaternary]. Such widespread use has led to the continuous discharge of these compounds into aquatic environments primarily via wastewater treatment plant effluent, with surface water concentrations being found to range from less than 1 to ~100 µg/L globally [@mohapatra2023quaternary]. Though the half-life of BAC in its aqueous form is estimated to be 5 to 7 days, its continuous input into freshwater systems creates pseudo-persistence and ongoing exposure of downstream aquatic biota [@marteinson2022increased; @deleo2020assessment]. 

```{r, echo=FALSE, message=FALSE}

library(here)
knitr::include_graphics(paste0(here(), "/images/wwtp_path.jpg"))

```

  Little is known about the effects of chronic BAC exposure in aquatic environments, particularly on lower-trophic-level organisms such as biofilm, phytoplankton, and zooplankton [@marteinson2022increased]. While some laboratory studies have investigated the acute toxicity of BAC on aquatic species – finding algae to be particularly sensitive [@hora2020increased; @jing2012quantitative] – these studies fail to capture the complexity of natural ecosystems and lack the ability to assess the indirect effects of a contaminant. 

  
```{r, echo=FALSE, message=FALSE}

library(here)
knitr::include_graphics(paste0(here(), "/images/bac_c12.png"))

```

  BAC is a cationic surfactant containing a positively charged ammonium headgroup bonded to an alkyl chain of varying lengths ranging from 8 to 18 carbons [@arnold2023quat]. It’s efficacy in breaking down bacterial cell walls is due to the electrostatic interactions between the positively charged quaternary nitrogen headgroup and negatively charged phospholipids and structural bacterial proteins of the cell membrane [@maillard2002bacterial; @gilbert2005cationic; @jennings2015quaternary]. A similar mode of action is seen against the protective lipid layer of viruses [@mohapatra2023quaternary]. Generally, antimicrobial activity increases with increasing alkyl chain length, with BAC C12 to C16 being the most effective [@gerba2015quaternary] The same properties that make BAC effective against bacteria and viruses also make it potentially toxic to aquatic biota, particularly lower-trophic-level organisms.

Considering the increased use and environmental release of antimicrobial products since the onset of COVID-19, it is essential to study their effects on aquatic organisms [@marteinson2022increased]. The abundance and diversity of lower-trophic-level organisms are integral to the structure and function of aquatic ecosystems. Biofilms and phytoplankton are important in dictating water quality, as they ensure the decomposition of organic matter and act as primary producers [@sentenac2022significance]. Meanwhile, zooplankton facilitate energy transfer up the food web and have impacts on lake biogeochemical processes and productivity [@pinel2021multi]. Zooplankton are known for their sensitivity to environmental change, including contaminants, and are useful indicators for detecting these disturbances [@pinel2021multi; @yan1996recovery]. Understanding the effects of BAC on aquatic biota, particularly lower-trophic-level organisms, is necessary to assess potential bottom-up effects this compound may have on aquatic ecosystems in Canada. 

### *Objective*

The objective of my research is to investigate the effects of chronic BAC exposure on biofilm, phytoplankton, and zooplankton communities in a northern Ontario lake. I predict that with increasing BAC concentration, biofilm and phytoplankton chlorophyll a will decrease, thus reducing the abundance and diversity of the zooplankton communities that rely on these food sources.

### *Methodology*

Beginning in June of 2023, an eleven-week mesocosm study was conducted at the International Institute for Sustainable Development – Experimental Lakes Area (IISD-ELA) to assess the effects of chronic BAC exposure on aquatic ecosystems using five environmentally relevant concentrations (nominal 20, 112, 632, 3556, and 20 000 ng/L; with three controls of 0 ng/L). Eight mesocosms – enclosures made of a floating, 2 m diameter ring with an impermeable polyethylene curtain sandbagged to the bottom of the lake – were deployed in the north basin of Lake 375. Enclosures were spiked weekly with a technical mixture of BAC homologs with varying alkyl chain lengths (C10 to C18) to mimic BAC’s pseudo-persistent behaviour and to ensure target concentrations were maintained. 

```{r, echo=FALSE, message=FALSE}

library(here)
knitr::include_graphics(paste0(here(), "/images/mesocosm_diagram.jpg"))

```

Colonization strips were installed in the photic zone of each enclosure and sampled for biofilm chlorophyll a and ash-free dry mass. Water samples from each enclosure were taken every 21 days to determine phytoplankton chlorophyll a and community assemblage [@havens2024chemical]. Zooplankton samples were collected using a Schindler-Patalas plankton trap lowered to 0.5 m and preserved in a 5% sugar-formalin solution at two- to three-week intervals [@prepas1978sugar]. In the laboratory, zooplankton samples were identified to species and life stage for abundance and diversity analyses. Multivariate analyses will be used to discern the drivers of shifts in community composition.

### *Data Analysis*



### *References*